# Preface

## R packages & versioning

The R packages listed below will be necessary at some point over the course of
this book. I recommend installing them all now. The block of code below is
designed to first check if each of the listed packages is already installed on
your computer. If any is missing, then an attempt is made to install it from
CRAN. Finally, all of the packages are loaded into the environment.

```{r package_loading, echo=TRUE, message=FALSE, warning=FALSE}
## Specify the packages you'll use in the script
packages <- c("tidyverse",
              "readxl",
              "zoo",
              "gridExtra",
              "R.matlab",
              "cowplot",
              "easystats",
              "circular",
              "splines",
              "MESS", ## area under curve
              "zoo" ## rolling means
)
## Now for each package listed, first check to see if the package is already
## installed. If it is installed, it's simply loaded. If not, it's downloaded 
## from CRAN and then installed and loaded.
package.check <- lapply(packages,
                        FUN = function(x) {
                          if (!require(x, character.only = TRUE)) {
                            install.packages(x, dependencies = TRUE)
                            library(x, character.only = TRUE)
                          }
                        }
)
```

I will use the `sessionInfo()` command to detail the specific versions of
packages I am using (along with other information about my R session). Please 
note that I am not suggesting you obtain exactly the same version of each 
package listed below. Instead, the information below is meant to help you assess
whether package versioning underlies any trouble you may encounter.

```{r package_versions, echo=FALSE}
print(sessionInfo())
```


## `%not_in%`

This guide will also rely on this handy function, which you should add to your
code:

``` {r not_in}
`%not_in%` <- Negate(`%in%`)
```

This simple operator allows you to determine if an element does not appear in a
target object.
