# Data wrangling

Once data have been spike sorted, we are ready to begin working in `R`. To get
to a point where meaningful preliminary plots can be produced, a few things need
to be addressed:

1) Labeling the time series of spike & photodiode data based on the stimulus
that appears on screen (i.e., matching the log file to the data file). This
includes labeling phases (like "blank", "stationary", and "moving") along with
experimental metadata such as SF, TF, and stimulus orientation (direction).

2) Re-organizing the spike & photodiode so that separate replicates of a
stimulus run are uniquely labelled and then arranged together.

3) Binning the data into 10- and 100-ms data sets, and then exporting CSVs of
the unbinned, 10-ms-binned, and 100-ms-binned data. This will be highly useful
for situations where you are handling multiple data files (e.g., from different
recording days), in which case it is likely that your machine will not be able
to store all objects in RAM.


## Packages

The packages listed below will be necessary at some point over the course of
this book. I recommend installing them all now. The block of code below is
designed to first check if each of the listed packages is already installed on
your computer. If any are missing, then an attempt is made to install it from
CRAN. Finally, all of the packages are loaded into the environment.

```{r package_loading, echo=TRUE, message=FALSE, warning=FALSE}
## Specify the packages you'll use in the script
packages <- c("tidyverse",
              "readxl",
              "zoo",
              "gridExtra",
              "R.matlab",
              "cowplot",
              "easystats",
              "circular",
              "splines",
              "MESS", ## area under curve
              "zoo" ## rolling means
)
## Now for each package listed, first check to see if the package is already
## installed. If it is installed, it's simply loaded. If not, it's downloaded 
## from CRAN and then installed and loaded.
package.check <- lapply(packages,
                        FUN = function(x) {
                          if (!require(x, character.only = TRUE)) {
                            install.packages(x, dependencies = TRUE)
                            library(x, character.only = TRUE)
                          }
                        }
)
```

I will use the `sessionInfo()` command to detail the specific versions of
packages I am using (along with other information about my R session).

```{r package_versions, echo=FALSE}
print(sessionInfo())
```


## `%not_in%`

This guide will also rely on this handy function, which you should add to your
code:

``` {r not_in}
`%not_in%` <- Negate(`%in%`)
```

This simple operator allows you to determine if an element does not appear in a
target object.


## Import example file and metadata

We will use a recently-collected data file and its corresponding metadata file
to showcase the fundamentals of wrangling ephys data into a more easily
plot-able format.

### Identify files to import
The following code is based on the assumptions that:
1) Your files are stored in a directory entitled `/data`
2) The basename of each file (i.e., the name of the file, excluding the file
extension) is identical for each set of spike sorted data and corresponding
metadata log file (e.g., `04132022_009m.mat` and `04132022_009m` have the same
basename, which is `04132022_009m`).

``` {r list_files}
## List all files of each file type
csv_files <-
  list.files("./data", pattern = ".csv",
             full.names = TRUE)
mat_files <-
  list.files("./data", pattern = ".mat",
             full.names = TRUE)

## Generate metadata tibbles for each file type
csv_file_info <-
  tibble(
    csv_files = csv_files,
    ## Extract the basename by removing the file extension
    basename =  basename(csv_files) %>% str_remove(".csv"),
    ## NOTE: PLEASE ADJUST THE FOLLOWING LINE TO BE ABLE TO EXCTRACT OUT THE
    ## DATE BASED ON YOUR NAMING CONVENTION
    basedate =  basename(csv_files) %>% str_sub(start = 1, end = 12)
  )
mat_file_info <-
  tibble(
    mat_files = mat_files,
    ## Extract the basename by removing the file extension
    basename =  basename(mat_files) %>% str_remove(".mat"),
    ## NOTE: AGAIN, PLEASE ADJUST THE FOLLOWING LINE TO BE ABLE TO EXCTRACT OUT
    ## THE DATE BASED ON YOUR NAMING CONVENTION
    basedate =  basename(mat_files) %>% str_sub(start = 1, end = 12)
  )

## Matchmake between .MAT data and .CSV log files
csv_mat_filejoin <-
  inner_join(csv_file_info, mat_file_info, by = "basename") %>%
  ## OPTIONAL STEP: remove any rows where either the .MAT or .CSV is missing
  drop_na()

## Store a vector of basenames in the environment. This will become useful later
base_names <- csv_mat_filejoin$basename

## Your end products from this code block should look something like:
csv_mat_filejoin

## and:
base_names
```

### Data import and preliminary labeling
We will now use the `R.matlab` package to import the `.mat` file into R and then
label the spike and photodiode time series based on the information in the
`.csv` log file

Because `.mat` files can be large, data import can take several minutes.

``` {r data_import}
## Tidy up how R has been using RAM by running garbage collection
gc()

## Set up empty vectors that will collect sets of replicates that we will be
## splitting up
metadata_sets <- NULL
meta_splits <- NULL
data_splits <- NULL
```
