# Direction tuning

This type of plot is so fundamental to the examination of LM and nBOR,
it deserves a chapter unto itself. The majority of the information
going into polar tuning plots is the same as those in previous
chapters. Unlike raster and mean spike rate plots, however, direction
tuning plots require deliberate choices from the investigator to
delineate a "baseline" spike rate to be distinguished from the spike
rate during the stimulus of interest (in our case, global motion
patterns). The conditions that define the baseline and motion epochs
require explicit definitions form the investigator.

Seeing the mean spike rate plot from the previous chapter will help
give context to some of these decisions **[INSERT IMAGE HERE]**.
Importantly, at the onset of the blank and the stationary phases, it
is common to observe an initial transient response -- a sharp increase
in the spike rate of the neuron -- followed by a return to a steady
state.

For our purposes, the baseline will be defined as the steady state
response during the stationary phase of the stimulus. We will collect
this steady state response rate across all stimulus conditions (i.e.,
varying combos of speed and direction), and will simply average all of
those response rates to attain our baseline rate (and its SEM). Please
note that it is up to the investigator to determine if this definition
is appropriate, especially if other stimulus presentations are being
used.

For the "motion epoch", I will use a few different definitions. Some
of these conventions are informed by previous work (e.g., Smyth et al.
2022), whereas others are just based on a rough approximation of what
may be appropriate for these data. The definitions of the motion
epochs will be:

1.  The entire 3-sec motion epoch
2.  The "initial transient" phase: 40-200ms after the onset of
    motion (as used in Smyth et al. 2022)
3.  The "steady state" phase: 1500-3000ms after the onset of motion,
    a.k.a., the second half of the motion stimulus (as used in Smyth
    et al. 2022)
4.  The first 500 ms of the motion phase. This is an arbitrary
    definition for demonstrative purposes only, but somewhat informed
    by observing the patterns in the mean spike rate plot

As we construct polar plots, two additional aspects will be included
in the plots:

1.  The "preferred direction" of the neuron, as defined by the vector
    sum method **[insert citation]**. This will be shown as a single 
    grey line in the plot that indicates the preferred direction. Bear
    in mind that this value may not always be meaningful, particularly
    in cases of multi-modal responses.
2.  The Sensitivity Index (SI) of the neuron, as defined by
    **[citation]**. This metric calculates the narrowness of tuning.
    An SI of 1 indicates strong response to a single direction whereas
    0 indicates similar response across all investigated directions.

## Data import and baseline rate measurement

We will use the 10-ms binned data for these examples. We'll read in the
file and then extract the baseline rate as defined above.

```{r polar_import_and_baseline}
## Read in the data
## Since there is only 1 example file, we'll simplify things by just
## subsetting bin10_filelist
bin10_data <-
  read_csv(bin10_filelist[1], show_col_types = FALSE) %>%
  as_tibble() %>%
  ## Again, we will set Speed as an ordered factor to help control plotting
  ## later on
  mutate(across(
    Speed,
    factor,
    levels = c("1024", "644", "407", "256", "128", "64", "32", "16", "8", "4")
  ))

## Extract all rows corresponding to our desired baseline epoch
baseline_df <-
  bin10_data %>%
  filter(Time_stand >= Blank_end + 0.5) %>% ## 0.5 sec after blank end
  filter(Time_stand <= Static_end - 0.05)   ## 0.05 sec before static end

## Compute the mean baseline
global_base <- mean(baseline_df$Spike_rate)
## Compute the SE
global_base_se <-
  sd(baseline_df$Spike_rate) / sqrt(length(baseline_df$Spike_rate)#/45
  )

## Construct a summary data frame
baseline_summary_df <-
  baseline_df %>%
  group_by(Speed) %>%
  summarize(
    speed_specific_baseline = mean(Spike_rate),
    speed_specific_baseline_se = sd(Spike_rate)/sqrt(length(Spike_rate)),
    global_baseline = global_base,
    global_baseline_se = global_base_se
  )

## This tibble contains speed-specific baselines (and SE) along with the global
## mean baseline (and SE)
baseline_summary_df
```

## Using the full 3-sec motion epoch

This will be referred to as the "naive" approach in the code below

```{r naive, fig.height=3, fig.width=10}
## global baseline values, full 3-sec motion period
naive_df <-
  bin10_data %>%
  filter(Time_stand > Static_end) %>%
  group_by(Speed, Direction, Replicate) %>%
  summarize(
    mean_spike_rate = mean(Spike_rate)
  )
naive_360 <-
  naive_df %>%
  filter(Direction == 0) %>%
  transmute(
    Speed = Speed,
    Direction = 360,
    Replicate = Replicate,
    mean_spike_rate = mean_spike_rate)
naive_vecsum_df <-
  naive_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  ## NOTE: FOLLOWING THE JNP AND CB PAPERS, WE ARE SUBTRACTING BASELINE HERE AND
  ## THEN IF ANY AVERAGED FIRING RATES ARE NEGATIVE, THEY ARE SHIFTED SO THE
  ## LOWEST ONE IS ZERO. THE GENERATED PLOTS STILL SHOW NON-BASELINE-SUBTRACTED
  ## FIRING RATES (AND BASELINE AS A RED RING), BUT COMPUTATION OF VECTOR SUM
  ## AND SI HAVE BEEN BASELINE SUBTRACTED (AND THE ENTIRE CURVE IS SHIFTED
  ## UPWARDS IF ANY PART IS NEGATIVE)
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      x = cos(Direction * pi / 180) * mean_spike_rate,
      y = sin(Direction * pi / 180) * mean_spike_rate,
      Speed = first(Speed)
  ) %>%
  map(summarise,
      x = mean(x),
      y = mean(y),
      Speed = first(Speed)) %>%
  map(transmute,
      vector_sum = (atan2(y, x) * 180 / pi) %% 360,
      Speed = first(Speed)
  ) %>%
  bind_rows()
naive_si_df<-
  naive_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      a = (sin(Direction * pi / 180) * mean_spike_rate),
      b = (cos(Direction * pi / 180) * mean_spike_rate),
      c = mean(mean_spike_rate),
      Speed = first(Speed)
  ) %>%
  map(summarise,
      a = mean(a),
      b = mean(b),
      c = mean(c),
      Speed = first(Speed)) %>%
  map(transmute,
      si = sqrt(a ^ 2 + b ^ 2) / c,
      Speed = first(Speed)
  ) %>%
  bind_rows()

naive_data <-
  naive_df %>%
  bind_rows(naive_360) %>%
  left_join(naive_vecsum_df, by = "Speed") %>%
  left_join(naive_si_df, by = "Speed") %>%
  drop_na(mean_spike_rate)
naive_max_y <- max(naive_data$mean_spike_rate)

naive <-
  naive_data %>%
  left_join(baseline_summary_df, by = "Speed") %>%
  ggplot(aes(x = Direction, y = mean_spike_rate)) +
  geom_ribbon(aes(
    x = Direction,
    ymin = global_baseline - global_baseline_se,
    ymax = global_baseline + global_baseline_se
  ),
  fill = "red") +
  stat_smooth(method = "glm", formula = y ~ ns(x,8), linewidth = 0.3) +
  geom_point() +
  geom_label(aes(label = round(si, 2) , x = 315, y = naive_max_y * 1.2),
             size = 3) +
  geom_vline(aes(xintercept = vector_sum), colour = "grey30",
             size = 0.75) +
  coord_polar(direction = 1, start = pi/2) +
  scale_x_continuous(
    breaks = c(0, 90, 180, 270),
    expand = c(0, 0),
    limits = c(0, 360)
  )+
  facet_grid(cols = vars(Speed)) +
  ggtitle("Full 3-sec motion") +
  ylab("Spike rate (spikes/sec)") +
  theme_minimal()

naive
```


## Using 40-200ms ("initial transient")

```{r cb_it, fig.height=3, fig.width=10}
## global baseline values, 40-200 msec motion period
cbit_df <-
  bin10_data %>%
  filter(Time_stand >= Static_end + 0.04) %>%
  filter(Time_stand <= Static_end + 0.2) %>%
  group_by(Speed, Direction, Replicate) %>%
  summarize(
    mean_spike_rate = mean(Spike_rate)
  )
cbit_360 <-
  cbit_df %>%
  filter(Direction == 0) %>%
  transmute(
    Speed = Speed,
    Direction = 360,
    Replicate = Replicate,
    mean_spike_rate = mean_spike_rate)
cbit_vecsum_df <-
  cbit_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      x = cos(Direction * pi / 180) * mean_spike_rate,
      y = sin(Direction * pi / 180) * mean_spike_rate,
      Speed = first(Speed)
  ) %>%
  map(summarise,
      x = mean(x),
      y = mean(y),
      Speed = first(Speed)) %>%
  map(transmute,
      vector_sum = (atan2(y, x) * 180 / pi) %% 360,
      Speed = first(Speed)
  ) %>%
  bind_rows()
cbit_si_df <-
  cbit_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      a = (sin(Direction * pi / 180) * mean_spike_rate),
      b = (cos(Direction * pi / 180) * mean_spike_rate),
      c = mean(mean_spike_rate),
      Speed = first(Speed)
  ) %>%
  map(summarise,
      a = mean(a),
      b = mean(b),
      c = mean(c),
      Speed = first(Speed)) %>%
  map(transmute,
      si = sqrt(a ^ 2 + b ^ 2) / c,
      Speed = first(Speed)
  ) %>%
  bind_rows()

cbit_data <-
  cbit_df %>%
  bind_rows(cbit_360) %>%
  left_join(cbit_vecsum_df, by = "Speed") %>%
  left_join(cbit_si_df, by = "Speed") %>%
  drop_na(mean_spike_rate)
cbit_max_y <- max(cbit_data$mean_spike_rate)

cb_it <-
  cbit_data %>%
  left_join(baseline_summary_df, by = "Speed") %>%
  ggplot(aes(x = Direction, y = mean_spike_rate)) +
  geom_ribbon(aes(
    x = Direction,
    ymin = global_baseline - global_baseline_se,
    ymax = global_baseline + global_baseline_se
  ),
  fill = "red") +
  stat_smooth(method = "glm", formula = y ~ ns(x,8), linewidth = 0.3) +
  geom_point() +
  geom_label(aes(label = round(si, 2) , x = 315, y = cbit_max_y * 1.2),
             size = 3) +
  geom_vline(aes(xintercept = vector_sum), colour = "grey30",
             size = 0.75) +
  coord_polar(direction = 1, start = pi/2) +
  scale_x_continuous(
    breaks = c(0, 90, 180, 270),
    expand = c(0, 0),
    limits = c(0, 360)
  )+
  facet_grid(cols = vars(Speed)) +
  ggtitle("40-200 ms (initial transient) motion") +
  ylab("Spike rate (spikes/sec)") +
  theme_minimal()

cb_it
```

## Using 1500-3000ms ("steady state")

```{r cb_ss, fig.height=3, fig.width=10}
## global baseline values, second half of motion period
cbss_df <-
  bin10_data %>%
  filter(Time_stand > Static_end + 1.5) %>%
  filter(Time_stand < 4.95) %>%
  group_by(Speed, Direction, Replicate) %>%
  summarize(
    mean_spike_rate = mean(Spike_rate)
  )
cbss_360 <-
  cbss_df %>%
  filter(Direction == 0) %>%
  transmute(
    Speed = Speed,
    Direction = 360,
    Replicate = Replicate,
    mean_spike_rate = mean_spike_rate)
cbss_vecsum_df <-
  cbss_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      x = cos(Direction * pi / 180) * mean_spike_rate,
      y = sin(Direction * pi / 180) * mean_spike_rate,
      Speed = first(Speed)
  ) %>%
  map(summarise,
      x = mean(x),
      y = mean(y),
      Speed = first(Speed)) %>%
  map(transmute,
      vector_sum = (atan2(y, x) * 180 / pi) %% 360,
      Speed = first(Speed)
  ) %>%
  bind_rows()
cbss_si_df <-
  cbss_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      a = (sin(Direction * pi / 180) * mean_spike_rate),
      b = (cos(Direction * pi / 180) * mean_spike_rate),
      c = mean(mean_spike_rate),
      Speed = first(Speed)
  ) %>%
  map(summarise,
      a = mean(a),
      b = mean(b),
      c = mean(c),
      Speed = first(Speed)) %>%
  map(transmute,
      si = sqrt(a ^ 2 + b ^ 2) / c,
      Speed = first(Speed)
  ) %>%
  bind_rows()

cbss_data <-
  cbss_df %>%
  bind_rows(cbss_360) %>%
  left_join(cbss_vecsum_df, by = "Speed") %>%
  left_join(cbss_si_df, by = "Speed") %>%
  drop_na(mean_spike_rate)
cbss_max_y <- max(cbss_data$mean_spike_rate)


cb_ss <-
  cbss_data %>%
  left_join(baseline_summary_df, by = "Speed") %>%
  ggplot(aes(x = Direction, y = mean_spike_rate)) +
  geom_ribbon(aes(
    x = Direction,
    ymin = global_baseline - global_baseline_se,
    ymax = global_baseline + global_baseline_se
  ),
  fill = "red") +
  stat_smooth(method = "glm", formula = y ~ ns(x,8), linewidth = 0.3) +
  geom_point() +
  geom_label(aes(label = round(si, 2) , x = 315, y = cbss_max_y * 1.2),
             size = 3) +
  geom_vline(aes(xintercept = vector_sum), colour = "grey30",
             size = 0.75) +
  coord_polar(direction = 1, start = pi/2) +
  scale_x_continuous(
    breaks = c(0, 90, 180, 270),
    expand = c(0, 0),
    limits = c(0, 360)
  )+
  facet_grid(cols = vars(Speed)) +
  ggtitle("1500-3000 ms (steady state) motion") +
  ylab("Spike rate (spikes/sec)") +
  theme_minimal()

cb_ss
```

## Using 0-500 ms (arbitrary epoch length)

```{r arb, fig.height=3, fig.width=10}
## global baseline values, 0-500 msec motion period
arb_df <-
  bin10_data %>%
  filter(Time_stand >= Static_end + 0.001) %>%
  filter(Time_stand <= Static_end + 0.500) %>%
  group_by(Speed, Direction, Replicate) %>%
  summarize(
    mean_spike_rate = mean(Spike_rate)
  )
arb_360 <-
  arb_df %>%
  filter(Direction == 0) %>%
  transmute(
    Speed = Speed,
    Direction = 360,
    Replicate = Replicate,
    mean_spike_rate = mean_spike_rate)
arb_vecsum_df <-
  arb_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      x = cos(Direction * pi / 180) * mean_spike_rate,
      y = sin(Direction * pi / 180) * mean_spike_rate,
      Speed = first(Speed)
  ) %>%
  map(summarise,
      x = mean(x),
      y = mean(y),
      Speed = first(Speed)) %>%
  map(transmute,
      vector_sum = (atan2(y, x) * 180 / pi) %% 360,
      Speed = first(Speed)
  ) %>%
  bind_rows()
arb_si_df <-
  arb_df %>%
  ungroup() %>%
  drop_na(mean_spike_rate) %>%
  group_split(Speed) %>%
  map(group_by, Direction) %>%
  map(summarize,
      mean_spike_rate = mean(mean_spike_rate) - global_base,
      Speed = first(Speed)) %>%
  map(mutate,
      mean_spike_rate =
        case_when(min(mean_spike_rate) < 0 ~ mean_spike_rate + abs(min(mean_spike_rate)),
                  TRUE ~ mean_spike_rate)) %>%
  map(transmute,
      a = (sin(Direction * pi / 180) * mean_spike_rate),
      b = (cos(Direction * pi / 180) * mean_spike_rate),
      c = mean(mean_spike_rate),
      Speed = first(Speed)
  ) %>%
  map(summarise,
      a = mean(a),
      b = mean(b),
      c = mean(c),
      Speed = first(Speed)) %>%
  map(transmute,
      si = sqrt(a ^ 2 + b ^ 2) / c,
      Speed = first(Speed)
  ) %>%
  bind_rows()

arb_data <-
  arb_df %>%
  bind_rows(arb_360) %>%
  left_join(arb_vecsum_df, by = "Speed") %>%
  left_join(arb_si_df, by = "Speed") %>%
  drop_na(mean_spike_rate)
arb_max_y <- max(arb_data$mean_spike_rate)

arb <-
  arb_data %>%
  left_join(baseline_summary_df, by = "Speed") %>%
  ggplot(aes(x = Direction, y = mean_spike_rate)) +
  geom_ribbon(aes(
    x = Direction,
    ymin = global_baseline - global_baseline_se,
    ymax = global_baseline + global_baseline_se
  ),
  fill = "red") +
  stat_smooth(method = "glm", formula = y ~ ns(x,8), linewidth = 0.3) +
  geom_point() +
  geom_label(aes(label = round(si, 2) , x = 315, y = arb_max_y * 1.2),
             size = 3) +
  geom_vline(aes(xintercept = vector_sum), colour = "grey30",
             size = 0.75) +
  coord_polar(direction = 1, start = pi/2) +
  scale_x_continuous(
    breaks = c(0, 90, 180, 270),
    expand = c(0, 0),
    limits = c(0, 360)
  )+
  facet_grid(cols = vars(Speed)) +
  ggtitle("0-500 msec motion") +
  theme_minimal()

arb
```

## Construct a multi-panel plot

We can use the handy `cowplot` package to construct a multi-panel plot.

```{r polar_cows, fig.height=8, fig.width=10}
cow_polar <-
  plot_grid(naive, cb_it, cb_ss, arb,
            ncol = 1)

cow_polar
```

This plot can also be written directly to `PDF`.
```{r polar_plot_pdf, eval=FALSE}
pdf(file =
      paste0("./path/to/directory/",
             bin10_basenames[i],
             "_polar_set.pdf"),
    width = 22, height = 17,
    pagecentre = TRUE, colormodel = "srgb")
plot(cow_polar)
dev.off()
```

